---
title: '6\. Worksheet: Among Site (Beta) Diversity -- Part 2'
author: "Madison Brown; Z620: Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

In this worksheet, we continue to explore concepts, statistics, and visualizations related to $\beta$-diversity.
Now that you know how to formally quantify $\beta$-diversity, we will learn how to test hypotheses about $\beta$-diversity using multivariate statistics. 

## Directions:
1. In the Markdown version of this document in your cloned repo, change "Student Name" on line 3 (above) with your name.
2. Complete as much of the worksheet as possible during class.
3. Use the handout as a guide; it contains a more complete description of data sets along with examples of proper scripting needed to carry out the exercises.
4. Answer questions in the worksheet.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio (color may vary if you changed the editor theme). 
5. Before you leave the classroom today, you should **push** this file to your GitHub repo, at whatever stage you are. This will enable you to pull your work onto your own computer.
6. When you have completed the worksheet, **Knit** the text and code into a single PDF file by pressing the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your Posit.cloud workspace: `/cloud/project/QB-2025/Week4-Beta/`
7. After Knitting, please submit the worksheet by making a **push** to your GitHub repo and then create a **pull request** via GitHub.
Your pull request should include this file (**6.BetaDiversity_2_Worksheet.Rmd**) 
with all code blocks filled out and questions answered) and the PDF output of `Knitr`  
(**6.BetaDiversity_2_Worksheet.pdf**).

The completed exercise is due on **Wednesday, February 12^th^, 2025 before 12:00 PM (noon)**.

## 1) R SETUP

Typically, the first thing you will do in either an R script or an RMarkdown file is setup your environment. 
This includes things such as setting the working directory and loading any packages that you will need.

In the R code chunk below, provide the code to:  

1. clear your R environment,
2. print your current working directory,
3. set your working directory to your `Week4-Beta/` folder. 
4. load the `vegan` R package (be sure to install if needed).  

```{r}
rm(list = ls())

getwd()

setwd("/cloud/project/QB2025_Brown/Week4-Beta")

package.list <- c('vegan', 'ade4', 'viridis', 'gplots', 'BiodiversityR', 'indicspecies')
library(vegan)
library(ade4)
library(viridis)
library(gplots)
library(indicspecies)



```

## 2) LOADING DATA
### Load dataset
In the R code chunk below, load the `doubs` dataset from the `ade4` package

```{r}
# note, pleae do not print the dataset when submitting

data(doubs)

fish <- doubs$fish
fish <- fish[-8,]

```
## 3) HYPOTHESIS TESTING

### A.  Multivariate Procedures for Categorical Designs
Earlier work done in the Doubs River suggested that the river has four distinct regions of habitat quality: the first region (sites 1-14) of "high quality"; the second (sites 15 - 19) and fourth (sites 26 - 30) of "moderate quality"; and the third  (sites 20 - 25) of "low quality".

In the code chunk below, test the hypothesis that fish community composition varies with river quality.

1. create a factor vector that categorizes habitat quality in the Doubs River,
2. use the multivariate analyses for categorical predictors to describe how fish community structure relates to habitat quality.

```{r}

quality <- c(rep("HQ", 13), rep("MQ", 5), rep("LQ", 6), rep("MQ", 5))

adonis2(fish ~ quality, method = "bray", permutations = 999)

indval <- multipatt(fish, cluster = quality, func = "IndVal.g" ,
                    control = how(nperm=999))
summary(indval)

fish.rel <- decostand(fish, method = "total")
phi <- multipatt(fish.rel, cluster = quality, func = "r.g",
                 control = how(nperm=999))
summary(phi)
  
```

***Question 1***: Based on the PERMANOVA, IndVal, and phi coefficient analyses, what did you learn about the relationship between habitat quality and the fish species composition?
Are the different analyses consistent with one another and do they agree with the visualizations (heat maps, cluster dendograms, ordinations) that you created?

> ***Answer 1***: Based on the three analyses ran, all resulted in small p-values (<0.05) and overall statistically significant findings. This implies that there is a relationship between habitat quality and fish composition. These models do not necessarily give insight into what the relationship is and how exactly fish species are impacted by habitat. Yet, it does provide a statistical basis that there is a relationship present. Normally, once you see that there is a relationship present, you would perform further analysis to determine the extent of the relationship. The analyses performed above are consistent and do agree with the visualizations that have been created. The visualizations that were created showed that there were differences in fish communities between different sites. This implies that when habitats are different, so are the fish communities.  


### B.  Multivariate Procedures for Continuous Designs

### i.  Mantel Test
In the R code chunk below, do the following:   

1. create distance matrices for both fish communities and environmental factors, and
2. use a Mantel test to test the hypothesis that fish assemblages are correlated with stream environmental variables.

```{r}
fish.dist <- vegdist(doubs$fish[-8, ], method = "bray")
env.dist <- vegdist(scale(doubs$env[-8,]),method = "euclid")

mantel(fish.dist,env.dist)

```

***Question 2***: What do the results from our Mantel test suggest about fish diversity and stream environmental conditions? 
How does this relate to your hypothesis about stream quality influencing fish communities?

> ***Answer 2***: The Mantel test resulted in an r value of 0.604 and a p-value of 0.001. An r value of 0.604 means that there is a moderate positive correlation between fish diversity and stream environmental conditions. Furthermore, as stream environmental conditions change, fish diversity changes as well. The small p-value indicates that this finding is statistically signifcant. This further supports the hypothesis that as stream quality changes, fish communities will change as well.   

### ii.  Constrained Ordination
In the R code chunk below, do the following:  

1. create an environmental matrix of the water chemistry data included in the `doubs` dataset using forward and reverse selection of variables,
2. conduct a redundancy analysis on the fish assemblages of the Doubs River,
3. use a permutation test to determine the significance of the constrained analysis,
4. use a permutation test to determine the correlation of each environmental factor on the constrained axes,
5. calculate the explained variation on the first and second constrained axes,
6. plot the constrained ordination results including labeled points for each site, and
7. add vectors that demonstrate the influence of each environmental factor the constrained ordination.   
```{r}
env.chem <- as.matrix(doubs$env[-8 , 5:11])

fish.db <- vegdist(fish, method = "bray", upper = TRUE, diag = TRUE)

doubs.dbrda <- dbrda(fish.db ~ ., as.data.frame(env.chem))
ordiplot(doubs.dbrda)

doubs.dbrda.mod0 <- dbrda(fish.db ~ 1, as.data.frame(env.chem))

ordiplot(doubs.dbrda.mod0)

doubs.dbrda.mod1 <- dbrda(fish.db ~ ., as.data.frame(env.chem))

doubs.dbrda <- ordiR2step(doubs.dbrda.mod0, doubs.dbrda.mod1, perm.max = 200)

doubs.dbrda$call
doubs.dbrda$anova
ordiplot(doubs.dbrda)

permutest(doubs.dbrda, permutations = 999)
envfit(doubs.dbrda, env.chem[,c(4,6,7)], perm = 999)

dbrda.explainvar1 <- round(doubs.dbrda$CCA$eig[1] /
                             sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) * 100
dbrda.explainvar2 <- round(doubs.dbrda$CCA$eig[2] /
                             sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) * 100

par(mar = c(5, 5, 4, 4) + 0.1)

plot(scores(doubs.dbrda, display = "wa"), xlim = c(-1.3, 1.1),
     ylim = c(-1.1, 2.7), xlab = paste("dbRDA 1 (", dbrda.explainvar1, "%)",
     sep = ""), ylab = paste("dbRDA 2 (", dbrda.explainvar2, "%)",
     sep = ""), pch = 16, cex = 2.0, type = "n", cex.lab = 1.5,
     cex.axis = 1.2, axes = FALSE)

axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)

points(scores(doubs.dbrda, display = "wa"),
       pch = 19, cex = 3, bg = "gray", col = "gray")
text(scores(doubs.dbrda, display = "wa"),
     labels = row.names(scores(doubs.dbrda, display = "wa")))

vectors <- scores(doubs.dbrda, display = "bp")
arrows(0, 0, vectors[,1], vectors[, 2],
       lwd = 2, lty = 1, length = 0.2, col = "red")
text(vectors[,1], vectors[, 2], pos = 3,
     labels = row.names(vectors))
axis(side = 3, lwd.ticks=2, cex.axis=1.2, las = 1, col = "red", lwd = 2.2,
     at = pretty(range(vectors[, 1])) * 2, labels = pretty(range(vectors[, 1])))
axis(side = 4, lwd.ticks=2, cex.axis=1.2, las = 1, col = "red", lwd = 2.2,
     at = pretty(range(vectors[, 2])) * 2, labels = pretty(range(vectors[, 2])))

```

***Question 3***: Based on the constrained ordination, what are the environmental variables (or groups of correlated variables) that seem to be contributing to variation in fish community structure?

> ***Answer 3***: Based on the constrained ordination model, it appears that nitrogen and oxygen contribute to variation in fish community. These two variables have the longest arrows, which implies they have a stronger impact on the community and explain more of the variation. Bdo also contributes to variation in the fish community, but the arrow is not as long. Therefore, it likely does not explain as much variance as Oxy and Nit does. 

### iii. Variation Partitioning
In the code chunk below,   

1. Create a matrix model of the selected environmental variables,
2. Create a matrix model of the selected PCNM axes,
3. Perform constrained and partial constrained ordinations using the spatial and environmental models you just created,
4. Test the significance of each of your constrained ordinations using permutation tests,
5. Partition the variation among sites into the relative importance of space, environment, spatially structured environment, and residuals,
6. Plot the variation partitioning output to visualize it.  
```{r}
doubs.dbrda$anova

env.mod <- model.matrix(~ oxy + bdo + nit, as.data.frame(env.chem))[,-1]

rs <- rowSums(fish)/sum(fish)

doubs.pcnmw <- pcnm(dist(doubs$xy[-8,]), w = rs, dist.ret = T)

doubs.pcnmw$values > 0

doubs.space <- as.data.frame(scores(doubs.pcnmw))
doubs.pcnm.mod0 <- dbrda(fish.db ~ 1, doubs.space)
doubs.pcnm.mod1 <- dbrda(fish.db ~ ., doubs.space)
step.pcnm <- ordiR2step(doubs.pcnm.mod0, doubs.pcnm.mod1, perm.max = 200)

plot(step.pcnm)

step.pcnm$anova

space.mod <- model.matrix(~ PCNM2 + PCNM3 + PCNM5 + PCNM1 +
                            PCNM13 + PCNM16 + PCNM6, doubs.space)[,-1]

doubs.total.env <- dbrda(fish.db ~ env.mod)
doubs.total.space <- dbrda(fish.db ~ space.mod)

doubs.env.cond.space <- dbrda(fish.db ~ env.mod + Condition(space.mod))
doubs.space.cond.env <- dbrda(fish.db ~ space.mod + Condition(env.mod))

permutest(doubs.env.cond.space, permutations = 999)
permutest(doubs.space.cond.env, permutations = 999)
permutest(doubs.total.env, permutations = 999)
permutest(doubs.total.space, permutations = 999)

doubs.varpart <- varpart(fish.db, env.mod, space.mod)
doubs.varpart

par(mar = c(2,2,2,2))
plot(doubs.varpart)
text(1, 0.25, "Space")
text(0, 0.25, "Env")
mtext("Variation Partitioning of\nDoubs Fish Diversity", side = 3, line = -3)
```

***Question 4***: Interpret the variation partitioning results.   

> ***Answer 4***: Based on the variation partitioning results, 13% of the variation in fish community structure can be explained solely by environmental factors. On the other hand, 24% of the variation in fish community structure can be explained solely by spatial position. Based on this information, when attempting to identify one singular variable that has a larger impact on fish community structure, spatial position has a stronger impact. When taking into account spatially structured environmental variation, the two variables combined explain 37% of the variance in fish community structure. That leaves approximately 26% of the variance unaccounted for. 

## SYNTHESIS

Load the dataset from that you and your partner are using for the team project. 
Use one of the hypothesis-testing tools introduced in the beta diversity module. 
Interpret the findings of your data with respect to principles of biodiversity. 

```{r}

fish_data <- read.csv("/cloud/project/QB2025_Brown/Fish_Dataset.csv")

fish_data$Latitude <- cut(fish_data$Latitude, breaks=c(20, 25, 30, 35, 40, 45, 50),
                         labels=c("20-25", "25-30", "30-35", "35-40", "40-45", "45-50"),
                         include.lowest=TRUE)

Fish_lat <- cbind(fish_data[ ,2] ,fish_data[ ,23:658])
Fish_lat <- data.frame(Latitude = fish_data$Latitude, Fish_lat)

result <- aggregate(. ~ Latitude, data=Fish_lat, sum, na.rm = TRUE)

result <- cbind(result[ ,1], result[,3:638])

new_result <- result[ ,2:637]

colnames(result)[1] <- "Latitude Group"

new_result <- result[ ,2:637]

fish.db1 <- vegdist(new_result, method = "bray", upper = TRUE, diag = TRUE)

Fishdbmatrix <- as.matrix(fish.db1)

order2 <- rev(rownames(Fishdbmatrix))

# Beginning of beta diversity week 2 (above is beta diversity week 1)

fish_datanew <- cbind(fish_data[, 2:3], fish_data[, 5], fish_data[, 7:9], fish_data[, 23:658])

colnames(fish_datanew)[3] = "Water_Temp"
#fish_datanew

water.mod <- model.matrix(~ Water_Temp + pH + Nitrate_ug_L + TotalPhosphorus_ug_L, as.data.frame(fish_datanew))[,-1]

only.species <- fish_datanew[, 7:642]

species.db <- vegdist(only.species, method = "bray", upper = TRUE, dig = TRUE)

water.matrix <- as.matrix(fish_data[, 5:9])

fishdbrda <- dbrda(species.db ~ ., as.data.frame(water.matrix))
ordiplot(fishdbrda)

```

> ***Synthesis Answer***: The ordination plot above indicates that water temperature and pH have a stronger impact on the fish communities. The long arrow lengths are what indicate a strong relationship. There are other water characteristics in this plot including conductivity and the concentration of nitrate and phosphorus. However, given that the length of those arrows are much shorter, that implies that they do not have as strong of an impact on the fish communities. The plot is somewhat difficult to interpret due to the vast number of plotted points. Further data organization will be necessary to try to condense some of the data sites, given the difficulty of plotting over 2000 sites.
